Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-17 E-Hypergraphs-LICS24.tex
completeness proof update.

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(e) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are lead to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ in any hypergraph $G$ and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{???} is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs}, hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

Precisely,  our soundness and completeness result is the statement that,  for any morphisms $f$ and $g$ in an appropriate free semilattice-enriched SMC,  $f = g$  if and only if there exists a sequence of DPO-rewrites --- each induced by a structural equality or the monoidal theory --- between their combinatorial representations. 
In the particular case of a monoidal signature including natural copy and delete maps,  we thus recover a sound and complete representation of semilattice-enriched Cartesian categories,  with the translation described earlier in the introduction justifying our claim of developing a mathematical theory of e-graphs.

@@

Although e-graphs were first developed in the 80's \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesizers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}.  We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  Some further avenues for investigation, including potential applications, are given in the conclusion of this paper.

Our use of string diagrams is also central to the intuition behind our work.  There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica_hierarchical_2023} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
Hierarchical (hyper-)graphs are a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
DAN: THIS NEEDS TO BE CAREFULLY EXPANDED TO NOT OFFEND POTENTIAL REVIEWERS.

@@

Figure \ref{fig:egraph-strings} displays the generators of this language and Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
The first four equations are those displayed in Definition \ref{def:def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
We use string diagrams informally in this paper, but it is intuitively clear that these diagrams are sound and complete with respect to their intended categorical semantics. 
Indeed, similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below

% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}

% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
% \[\begin{tikzcd}
% 	\mathcal L && {i+j} \\
% 	\\
% 	{\mathcal G^{\urcorner}} && {\mathcal{L}^{\bot}} \\
% 	\\
% 	&& {n+m}
% 	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
% 	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
% 	\arrow["f"', from=1-1, to=3-1]
% 	\arrow["{[a_1,a_2]}"', from=1-3, to=1-1]
% 	\arrow["{[c_1,c_2]}", from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% \end{tikzcd}\]

\[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_boundary_complement}
    }
\]

is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is a monogamous cospan. 
\end{definition}
%\begin{remark}
%    Note how the requirement for $i \xrightarrow{} L \xleftarrow{} j$ and $n \xrightarrow{} G \xleftarrow{} m$ being monogamous, i.e. morphisms in $\MdaCospans$, is a part of the above definition.
%\end{remark}
Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.

@@

The appropriate notion of \textit{match} for our hierarchical setting additionally requires that its image is a \textit{down-closed} graph,  as defined below. 
This ensures the result of a rewrite is a well-typed MDA e-hypergraph.

\begin{definition}[Down-closed subgraph]
%A \textit{subgraph} of $\mathcal G $ an e-hypergraph $\mathcal H$ 
%is an e-hypergraph whose edge and vertex sets are subsets of those of $\mathcal H$, 
%and whose remaining equipment is a restriction of that of $\mathcal H$ to those sets. 
    We call  a sub-e-hypergraph $\mathcal G $ of $\mathcal H$ \emph{down-closed} if for all $e \in E_{\mathcal{G}}$,   all children of $e$ are also in $\mathcal{G}$.
\end{definition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces are given by the following:
		\begin{align*}
		&n'' = (i'' \setminus i) + l\\
		&k'' = (j'' \setminus j) + r\\
		&n'' \to \mathcal{H} = [f_1,f_2] \text{ where }\\
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H}\\ 
		&f_2 : l \to \mathcal{C} \to \mathcal{H} 
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
Since both $f_1$ and $f_2$ are monos with disjoint image (?), their copairing is also a mono. 
Similarly for $k'' \to \mathcal{H}$.
CHRIS: FIX (4) ABOVE:

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
%The conditions above, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
%\[
%	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
%\]
%We use 
With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense.

@@

This normal form can be used to prove the main theorem of the paper: the (full) completeness of our combinatorial representation of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$. 
\begin{theorem}[Full completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $(*)$ $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA cospans which is given by the restriction of $\llbracket - \rrbracket$ to $\textbf{SMT}(\Sigma, \mathcal E)$ as a subcategory of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$ (note that cospans embed in extended cospans when the internal interface is isomorphic to the external interface); see Theorem 35 of ~\cite{bonchi_string_2022-2}.

To see that the functor is faithful,  we prove that $f \neq g$ in $\textbf{SMT}^+(\Sigma, \mathcal E)$ implies $\llbracket f \rrbracket \neq \llbracket g \rrbracket$.  First,  calculate weak normal form $\Sigma^+$-terms of $f$ and $g$ as $f_1 + \ldots + f_n$ and $g_1 + \ldots + g_m$,  respectively.  Applying the functor $\llbracket - \rrbracket$,  we see that 
\begin{align*}
	\llbracket f \rrbracket = \llbracket f_1 + \ldots + f_n \rrbracket = \llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket \\
	\llbracket g \rrbracket = \llbracket g_1 + \ldots + g_m \rrbracket = \llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket ~ . 
\end{align*}
In particular,  note that $\llbracket - \rrbracket$ preserves weak normal forms.  Therefore $\llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket$ and $\llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket$ are weak normal forms.  Thus if $n \neq m$ we have that $\llbracket f \rrbracket \neq \llbracket g \rrbracket $.  Otherwise,  if $n = m$,  we have that there for every permutation $\sigma$ there exists an index $i$ such that $f_i \neq g_{\sigma(i)}$,  and thus by $(*)$ that $\llbracket f_i \rrbracket \neq \llbracket g_{\sigma(i)} \rrbracket $.  This implies that $\llbracket f \rrbracket \neq \llbracket g \rrbracket $.

To see that the functor is full,  consider a cospan $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ is a plain MDA cospan.  Thus,  by $(*)$ we can find morphisms $g_i$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we have a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}

@@

In each case above,  we aim to combine our combinatorial representation with existing combinatorial representations of the relevant structure.  Indeed,  the hypergraph representation and DPO-rewriting theory of each case is already well-studied independently of any e-graph structure \cite{ghica_rewriting_2023,cite,cite,cite}. % traced comonoid
%The hierarchical ``e-box'' structure within string diagrams and their associated e-hypergraphs (categorically modelled by semilattice enrichment) is essentially orthogonal to these developments. Combining the two, to give notions of e-hypergraphs specifically suited to these particular monoidal theories should be conceptually straightforward.  However, as we have seen in this paper, the technical details can be delicate. 
Finally,  a term calculus (e-terms) can be reverse engineered from the string-diagram notation~\cite{ghica_structural_nominal, Heijltjes:FMCII}. Although term calculi are less important in the algorithmics of optimisation, they still play a role in specification and in aiding readability.

>>>>

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(e) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are lead to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022} is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs}, hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

Precisely,  our soundness and completeness result is the statement that morphisms in an appropriate free semilattice-enriched SMC are equal  if and only if there exists a sequence of DPO-rewrites --- each induced by a structural equality or the monoidal theory --- between their combinatorial representations. 
In the particular case of a monoidal signature including natural copy and delete maps,  we thus recover a sound and complete representation of semilattice-enriched Cartesian categories,  with the translation described earlier in the introduction justifying our claim of developing a mathematical theory of e-graphs.

@@

% E-graphs
Although e-graphs were first developed in the 80's \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesizers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}.  We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  Some further avenues for investigation, including potential applications, are given in the conclusion of this paper.

% String diagrams
Our use of string diagrams is also central to the intuition behind our work \cite{Selinger_2010, joyal_geometry_1991}.  There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica_hierarchical_2023} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
% Hierarchical hypergraphs
Hierarchical (hyper-)graphs are also a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
% String diagram rewrites
Our work builds heavily on string diagram rewrite theory as developed in \cite{bonchi_string_2022,bonchi_string_2022-1, bonchi_string_2022-2}.

@@

Figure \ref{fig:egraph-strings} displays the generators of this language and Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
We use string diagrams informally in this paper, but it is intuitively clear that these diagrams are sound and complete with respect to their intended categorical semantics. 
Indeed, similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
% \[\begin{tikzcd}
% 	\mathcal L && {i+j} \\
% 	\\
% 	{\mathcal G^{\urcorner}} && {\mathcal{L}^{\bot}} \\
% 	\\
% 	&& {n+m}
% 	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
% 	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
% 	\arrow["f"', from=1-1, to=3-1]
% 	\arrow["{[a_1,a_2]}"', from=1-3, to=1-1]
% 	\arrow["{[c_1,c_2]}", from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% \end{tikzcd}\]
\[
    \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_boundary_complement}
    }
\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is a monogamous cospan. 
\end{definition}
%\begin{remark}
%    Note how the requirement for $i \xrightarrow{} L \xleftarrow{} j$ and $n \xrightarrow{} G \xleftarrow{} m$ being monogamous, i.e. morphisms in $\MdaCospans$, is a part of the above definition.
%\end{remark}
Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.

@@

The appropriate notion of \textit{match} for our hierarchical setting additionally requires that its image is a \textit{down-closed} graph,  as defined below. 
\begin{definition}[Down-closed subgraph]
%A \textit{subgraph} of $\mathcal G $ an e-hypergraph $\mathcal H$ 
%is an e-hypergraph whose edge and vertex sets are subsets of those of $\mathcal H$, 
%and whose remaining equipment is a restriction of that of $\mathcal H$ to those sets. 
    We call  a sub-e-hypergraph $\mathcal G $ of $\mathcal H$ \emph{down-closed} if for all $e \in E_{\mathcal{G}}$,   all children of $e$ are also in $\mathcal{G}$.
\end{definition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
        \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        \item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
		\begin{align*}
		&f_1 : i'' \setminus i \to \mathcal{R} \to \mathcal{H} & \\ 
		&f_2 : n' \setminus (i' \setminus i) \to \mathcal{C} \to \mathcal{H} & \\
		\end{align*}
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]

$[f_1,f_2]: n'' \to \mathcal{H} $
Since both $f_1$ and $f_2$ are monos with disjoint image (?), their copairing is also a mono.  Note $f_1$ is the obvious restriction,  $f_2$ is defined as $g_2$ in boundary complement. 
CHRIS: FIX (4) ABOVE:

    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

With our modified definition of extended DPOI rewriting in hand,  we proceed to prove that our combinatorial representation of semilattice-enriched SMCs is sound and complete in an appropriate sense.

@@

This normal form can be used to prove the main theorem of the paper: the (full) completeness of our combinatorial representation of $\textbf{SMT}^{+}(\Sigma, \mathcal {E} )$. 
\begin{theorem}[Full completeness]
We have the following equivalence of categories
\[
	\textbf{SMT}^{+}(\Sigma, \mathcal {E} ) \cong \WellTypedMdaEcospans / \mathcal{S,E}~.
\]
\end{theorem}
\begin{proof}
The equivalence is given by the functor $\llbracket - \rrbracket$.  First, note that the functor is surjective on objects,  so it suffices to prove the functor is both full and faithful.  In each case,  the result follows by a normal form argument in combination with relying on the equivalence $(*)$ $\textbf{SMT}(\Sigma, \mathcal E) \cong \textbf{MHypI}(\Sigma)/\mathcal{E}$ of symmetric monoidal theories and (appropriately quotiented) plain MDA cospans which is given by the obvious restriction of $\llbracket - \rrbracket$ to $\textbf{SMT}(\Sigma, \mathcal E)$; see Theorem 35 of ~\cite{bonchi_string_2022-2}.

To see that the functor is faithful,  we prove that $f \neq g$ in $\textbf{SMT}^+(\Sigma, \mathcal E)$ implies $\llbracket f \rrbracket \neq \llbracket g \rrbracket$.  First,  calculate weak normal form $\Sigma^+$-terms of $f$ and $g$ as $f_1 + \ldots + f_n$ and $g_1 + \ldots + g_m$,  respectively.  Applying the functor $\llbracket - \rrbracket$,  we thus have that
\[
	\llbracket f \rrbracket = \llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket \qquad
	\llbracket g \rrbracket = \llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket ~ . 
\]
Observe  that therefore $\llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket$ and $\llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket$ are weak normal forms.  We conisder two cases: where $n=m$ and where $n\neq m$.  In the first case,  it is immediate that $\llbracket f \rrbracket \neq \llbracket g \rrbracket $,  since their weak normal forms cannot be equal.  Otherwise,  if $n = m$,  and $\llbracket f \rrbracket = \llbracket g \rrbracket$,  we have that there exists a permutation $\sigma$ such that for every index $i$ we have $\llbracket f_i \rrbracket \neq \llbracket g_{\sigma(i)} \rrbracket $,  and thus by  $(*)$ that $f_i = g_{\sigma(i)}$,  implying that $f = g$ which is a contradiction.

To see that the functor is full,  consider a morphism $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ contains no hierarchical edges.  Thus,  by $(*)$ we can find morphisms $g_i$ in $\textbf{SMT}(\Sigma, \mathcal E)$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we can construct a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}

@@

In each case above,  we aim to combine our combinatorial representation with existing combinatorial representations of the relevant structure.  Indeed,  the hypergraph representation and DPO-rewriting theory of each case is already well-studied independently of any e-graph structure \cite{ghica_rewriting_2023,alvarez-picallo-functorial_2021}. % traced comonoid
%The hierarchical ``e-box'' structure within string diagrams and their associated e-hypergraphs (categorically modelled by semilattice enrichment) is essentially orthogonal to these developments. Combining the two, to give notions of e-hypergraphs specifically suited to these particular monoidal theories should be conceptually straightforward.  However, as we have seen in this paper, the technical details can be delicate. 
Finally,  a term calculus (e-terms) can be reverse engineered from the string-diagram notation~\cite{ghica_structural_nominal, Heijltjes:FMCII}. Although term calculi are less important in the algorithmics of optimisation, they still play a role in specification and in aiding readability.
